---
title: "Zona Incerta Regulation of Motivation - Dissertation Summary"
author: "Laura Korobkova"
date: '2025-11-02'
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

# Load all required libraries
library(ggplot2)
library(ggpattern)
library(ggpubr)
library(dplyr)
library(tidyr)
library(stringr)
library(grid)
library(VennDiagram)
library(kableExtra)
library(lme4, lib.loc = "/Library/Frameworks/R.framework/Versions/4.1/Resources/library")

# Define consistent color palette for the entire dissertation
COLORS <- list(
  # Treatment colors (blue-orange safe for colorblind)
  vehicle = "#0072B2",      # Blue for vehicle
  cno = "#E69F00",          # Orange for CNO

  # Group colors
  control = "#0072B2",      # Blue for controls
  gi_dreadd = "#D55E00",    # Vermillion for GI (inhibitory DREADD)
  gq_dreadd = "#009E73",    # Bluish green for GQ (excitatory DREADD)
  gfp = "#999999",          # Gray for GFP controls

  # Experiment-specific colors
  rewarded = "#009E73",     # Bluish green for rewarded
  no_reward = "#CC79A7",    # Reddish purple for non-rewarded
  stress = "#F0E442",       # Yellow for stress conditions
  fr1 = "#56B4E9",          # Sky blue

  # Stimulation colors
  no_stim = "#BFBFBF",       # Light gray                                       
  light_stim = "#0072B2",                                                       
  random_stim = "#404040",  # Dark gray for random stimulation

  # Sex colors (avoiding red-green)
  male = "#0072B2",         # Blue for males
  female = "#CC79A7"        # Reddish purple for females
)

# Custom theme for all plots (publication-ready)
theme_zi <- function() {                                                      
    theme_classic() +                                                           
      theme(                                                                    
        text = element_text(size = 26, family = "Arial"),                       
        plot.title = element_text(size = 18, face = "bold", hjust = 0.5),       
        axis.title = element_text(size = 16, face = "bold"),
        axis.text = element_text(size = 14),
        axis.text.x = element_text(angle = 0, hjust = 1),
        legend.position = "right",
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16, face = "bold"),
        legend.key.size = unit(1.2, "cm"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.text = element_text(size = 15, face = "bold"),
        strip.background = element_rect(fill = "gray95", color = NA)
      )
  }

# Define custom operator
'%!in%' <- function(x,y)!('%in%'(x,y))
```

# 1. GI DREADD in ZI: Motivation Experiments

## 1.1 Data Loading and Preprocessing

```{r h8090-data-prep}
# Load metadata and exclusion criteria
meta <- read.csv("data/meta_gi.csv")
excl <- read.csv("data/excluded_animals.csv")
breakpoint <- read.csv("data/breakpoint_gi.csv")

# Define comparison groups
group_comp <- list(
  c("cont_cno", "cont_veh"),
  c("GI_veh", "GI_cno"),
  c("cont_cno", "GI_cno"),
  c("cont_veh", "GI_veh")
)
```


#Progressive Ratio (PR) Test

```{r pr-test-analysis}
# Load and prepare PR data
pr_h80h90 <- read.csv("data/pr_gi.csv")
pr_h80h90 <- pr_h80h90[pr_h80h90$Animal %!in% excl$Animal,]
pr_h80h90 <- pr_h80h90[pr_h80h90$Animal != "H92.3",]
pr_h80h90 <- merge(pr_h80h90, breakpoint)
pr_h80h90$groups <- paste(pr_h80h90$group, pr_h80h90$treatment, sep="_")
pr_h80h90$groups <- factor(pr_h80h90$groups, levels = c("cont_veh", "cont_cno", "GI_veh", "GI_cno"))

knitr::kable(
  table(pr_h80h90[pr_h80h90$treatment=="cno",]$group,
        pr_h80h90[pr_h80h90$treatment=="cno",]$sex),
  caption = "Sample sizes by group and sex (CNO treatment)"
)

# Breakpoint analysis
gi_pr = ggplot(pr_h80h90[pr_h80h90$cohort %in% c("h90", "h80"),],
         aes(x = groups, y = breakingpoint)) +
    geom_boxplot(aes(fill = treatment), alpha = 0.7, outlier.shape = NA) +
    geom_boxplot_pattern(aes(pattern = group), alpha = 0.3) +
    geom_jitter(aes(color = sex), width = 0.1, size = 2.5, alpha = 0.8) +
    labs(
      title = "Progressive Ratio: GI DREADD Effect on Motivation",
      x = "Group and Treatment",
      y = "Breakpoint"
    ) +
    scale_fill_manual(
      name = "Treatment",
      values = c("cno" = COLORS$cno, "veh" = COLORS$vehicle)
    ) +
    scale_color_manual(
      name = "Sex",
      values = c("male" = COLORS$male, "female" = COLORS$female)
    ) +
    stat_compare_means(
      method = "wilcox.test",
      paired = FALSE,
      comparisons = group_comp,
      label = "p.signif",
      size = 7
    ) + theme(axis.text.x = element_text(face="bold", angle=0))+ theme_zi()

ggsave("/Users/laura/Desktop/GRADSCHOOL/DIAS/manuscripts_grants/ZI_MANUSCRIPT/FIGURES/R_gi_pr.pdf", plot = gi_pr, width = 11.5, 
       height=7.5,  units = "in", device = cairo_pdf, dpi = 300)

wilcox.test(pr_h80h90[pr_h80h90$cohort %in% c("h90", "h80") & pr_h80h90$groups=="GI_veh",]$breakingpoint, pr_h80h90[pr_h80h90$cohort %in% c("h90", "h80") & pr_h80h90$groups=="GI_cno",]$breakingpoint)

table(pr_h80h90[pr_h80h90$cohort %!in% c("zi-vp") & pr_h80h90$treatment=="veh",]$sex,pr_h80h90[pr_h80h90$cohort %!in% c("zi-vp") & pr_h80h90$treatment=="veh",]$group )

# Press rate analysis
ggplot(pr_h80h90, aes(x = groups, y = breakingpoint))  +
    geom_boxplot(aes(fill = treatment), alpha = 0.7, outlier.shape = NA) +
    geom_boxplot_pattern(aes(pattern = group), alpha = 0.3) +
    geom_jitter(width = 0.1, color = "black", size = 2.5, alpha = 0.6) +
    labs(
        title = "Progressive Ratio: GI DREADD vs GFP Control",
        x = "Group and Treatment",
        y = "Breakpoint"
    ) +
    scale_fill_manual(
        name = "Treatment",
        values = c("cno" = COLORS$cno, "veh" = COLORS$vehicle)
    ) +
    stat_compare_means(
        method = "wilcox.test",
        paired = FALSE,
        comparisons = group_comp,
        label = "p.signif",
        size = 5
    ) +
    theme_zi()

##ANOVA used in the paper 
model_gi_pr <- lmer(breakingpoint ~ treatment * group * sex + (1 | Animal), data =pr_h80h90)
anova(model_gi_pr)


# Statistical analysis
pr_anova <- aov(breakingpoint ~ treatment*group * sex , data = pr_h80h90[pr_h80h90$cohort %!in% c("zi-vp"),])
summary(pr_anova)

data_sub <- pr_h80h90[pr_h80h90$cohort %!in% c("zi-vp"), ]

library(effsize)

cont <- pr_h80h90[pr_h80h90$cohort %in% c("h90", "h80") & pr_h80h90$groups ==
  "cont_veh", ]$breakingpoint
  gi <- pr_h80h90[pr_h80h90$cohort %in% c("h90", "h80") & pr_h80h90$groups ==
  "GI_cno", ]$breakingpoint
  
  cohen.d(cont, gi)
  
```

**Key Finding:** Progressive ratio test indicates decreased motivation when ZI is inhibited `

#Open Field Test (Inhibition)

```{r open-field-analysis}
# Load and prepare open field data
open_field_h80h90 <- read.csv("data/open_field_gi.csv")
open_field_h80h90 <- open_field_h80h90[open_field_h80h90$Animal %!in% excl$Animal,]
open_field_h80h90 <- open_field_h80h90[open_field_h80h90$Animal != "H92.3",]
open_field_h80h90 <- merge(open_field_h80h90, meta)
open_field_h80h90$groups <- paste(open_field_h80h90$group, open_field_h80h90$treatment, sep="_")
open_field_h80h90$groups <- factor(open_field_h80h90$groups, levels = c("cont_veh", "cont_cno", "GI_veh", "GI_cno"))

table(open_field_h80h90$groups, open_field_h80h90$sex)

anova_open_gi = aov(Distance~ group*treatment*sex, data = open_field_h80h90[open_field_h80h90$excl!="excl" & open_field_h80h90$Distance<15000,])
summary(anova_open_gi)

ggplot(open_field_h80h90[open_field_h80h90$excl!="excl" & open_field_h80h90$Distance<15000,], aes(x = groups, y = Distance)) +
    geom_boxplot(aes(fill = treatment), alpha = 0.7, outlier.shape = NA) +
    geom_boxplot_pattern(aes(pattern = group), alpha = 0.3) +
    labs(
        title = "Open Field: Inhibitory DREADD Effect on Locomotion",
        x = "Group and Treatment",
        y = "Distance Traveled (cm)"
    ) +
    scale_fill_manual(
        name = "Treatment",
        values = c("cno" = COLORS$cno, "veh" = COLORS$vehicle)
    ) +
    stat_compare_means(
        method = "wilcox.test",
        paired = FALSE,
        comparisons = group_comp,
        label = "p.signif",
        size = 7
    ) +
    theme_zi() #+
geom_jitter(width = 0.1, color = "black", size = 2.5, alpha = 0.6) 


# By group faceted plot
open_field_h80h90$treatment = factor(open_field_h80h90$treatment, level = c("veh", "cno"))
ggplot(open_field_h80h90, aes(x = treatment, y = Distance)) +
  geom_boxplot(aes(fill = treatment), alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.1, color = "black", size = 2.5, alpha = 0.6) +
  labs(
    title = "Open Field: Distance by Group",
    x = "Treatment",
    y = "Distance Traveled (cm)"
  ) +
  scale_fill_manual(
    name = "Treatment",
    values = c("cno" = COLORS$cno, "veh" = COLORS$vehicle)
  ) +
  stat_compare_means(
    method = "wilcox.test",
    paired = FALSE,
    label = "p.signif",
    size = 7
  ) +
  facet_wrap(~group) +
  theme_zi()
```
**Key Finding:** Open field test shows that the ZI inhibition doesn't effect the locomotion.

##Memory Test 

```{r memory-test-analysis}
# Load and prepare memory test data
h90_memory <- read.csv("data/prob_memory_gi.csv")
meta_h90 <- read.csv("data/meta_gi_memory.csv")
h90_memory <- merge(h90_memory, meta_h90)
h90_memory$treatment <- factor(h90_memory$treatment, levels = c("veh", "cno"))

h90_memory$group_treat = paste(h90_memory$group, "_", h90_memory$treatment)

h90_memory$group_treat = factor(h90_memory$group_treat, levels = c("cont _ veh", "cont _ cno", "GI _ veh", "GI _ cno"))
comp_mem = list(
  c("cont _ veh", "cont _ cno"),
  c("cont _ veh", "GI _ veh"),
  c("cont _ cno", "GI _ veh"),
  c("GI _ veh", "GI _ cno")
)
  
  
# Memory test plot
memory_h90_plot = ggplot(h90_memory, aes(x = group_treat, y = lev1_over_lev2)) +
  geom_boxplot(aes(fill = treatment), alpha = 0.7, outlier.shape = NA) +
  labs(
    title = "No effect of GABAergic ZI inhibiton on Memory",
    x = "Treatment",
    y = "Correct Lever Index"
  ) +
  scale_fill_manual(
    name = "Treatment",
    values = c("cno" = COLORS$cno, "veh" = COLORS$vehicle)
  ) +
  stat_compare_means(
    method = "wilcox.test",
    paired = FALSE,
    label = "p.signif",
    size = 7, comparison = comp_mem
  ) + geom_boxplot_pattern(aes(pattern = group), alpha = 0.3) +theme_zi() # geom_jitter(width = 0.1, color = "black", size = 2.5, alpha = 0.6) 

ggsave("/Users/laura/Desktop/GRADSCHOOL/DIAS/manuscripts_grants/ZI_MANUSCRIPT/FIGURES/R_gi_memory.pdf", plot = memory_h90_plot, width = 11.5, 
       height=7.5,  units = "in", device = cairo_pdf, dpi = 300)



gi_memory_anova = aov(lev1_over_lev2 ~ treatment*group, data = h90_memory)
summary(gi_memory_anova)
```

**Key Finding:** ZI inhibition doesn't affect memory.

---

# 2. GQ DREADD in ZI: Excitatory Manipulation

## GQ DREADD Progressive Ratio

```{r gq-pr-analysis}

hcohorts_gq_pr = read.csv("data/pr_gq.csv")

hcohorts_gq_pr = merge(hcohorts_gq_pr, breakpoint)

hcohorts_gq_pr$group_treat <- factor(hcohorts_gq_pr$group_treat,
  levels = c("cont_veh", "cont_cno", "GQ_veh", "GQ_cno"))

group_comp_gq_cont <- list(
  c("cont_veh", "cont_cno"),
  c("cont_veh", "GQ_veh"),
  c("GQ_veh", "GQ_cno"),
  c("cont_veh", "GQ_cno"), c("cont_cno", "GQ_cno")
)

# GQ PR plot
gq_pr = ggplot(hcohorts_gq_pr, aes(x = group_treat, y = breakingpoint)) +
  geom_boxplot(aes(fill = treatment), alpha = 0.7, outlier.shape = NA) +
  geom_boxplot_pattern(aes(pattern = group), alpha = 0.3) +geom_jitter(aes(color = sex), width = 0.1, size = 2.5, alpha = 0.8) +
    labs(
        title = "Progressive Ratio: GQ DREADD Effect on Motivation",
        x = "Group and Treatment",
        y = "Breakpoint"
    ) +
    scale_fill_manual(
        name = "Treatment",
        values = c("cno" = COLORS$cno, "veh" = COLORS$vehicle)
    ) +
    scale_color_manual(
        name = "Sex",
        values = c("male" = COLORS$male, "fem" = COLORS$fem)
    ) +
    stat_compare_means(
        method = "wilcox.test",
        paired = FALSE,
        comparisons = group_comp_gq_cont,
        label = "p.signif",
        size = 7
    ) +
    theme_zi()

ggsave("/Users/laura/Desktop/GRADSCHOOL/DIAS/manuscripts_grants/ZI_MANUSCRIPT/FIGURES/R_gq_pr_final.pdf", plot = gq_pr, width = 11.5, 
       height=7.5,  units = "in", device = cairo_pdf, dpi = 300)


gq_pr_anova = aov(breakingpoint~group*treatment*sex, data = hcohorts_gq_pr)
summary(gq_pr_anova)

##USED in paper
model_gq <- lmer(breakingpoint ~ treatment * group*sex + (1 | animal), data = hcohorts_gq_pr)
anova(model_gq)

table(hcohorts_gq_pr[hcohorts_gq_pr$treat == "veh", ]$group, hcohorts_gq_pr[hcohorts_gq_pr$treat == "veh", ]$sex)

wilcox.test(hcohorts_gq_pr[hcohorts_gq_pr$group_treat=="GQ_veh",]$breakingpoint, hcohorts_gq_pr[hcohorts_gq_pr$group_treat=="GQ_cno",]$breakingpoint, paired=FALSE)

cont <- hcohorts_gq_pr[hcohorts_gq_pr$treatment=="cno" & hcohorts_gq_pr$group=="hm3dgq",]$breakingpoint

gq <- hcohorts_gq_pr[hcohorts_gq_pr$treatment=="veh" & hcohorts_gq_pr$group=="hm3dgq",]$breakingpoint

library(effsize) 

cohen.d(cont, gq)

```


## GQ Probe Test

```{r gq-probe-test}
# Load GQ probe test data
h_gq_prob_test <- read.csv("data/prob_test_gq.csv")
h_meta <- read.csv("data/meta_gq_gi.csv")
h_gq_prob_test <- merge(h_gq_prob_test, h_meta)

# GQ probe plot with individual trajectories
ggplot(
  h_gq_prob_test[h_gq_prob_test$animal %!in% c("18.2", "H112.2") &
                 h_gq_prob_test$cort != "cort",],
  aes(x = treatment, y = lever_pressed)
) +
  geom_boxplot(aes(fill = treatment), alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.1, size = 2.5, alpha = 0.6) +
  geom_line(aes(group = animal), alpha = 0.3) +
  labs(
    title = "GQ DREADD Probe Test: Lever Pressing Across Days",
    x = "Treatment",
    y = "Lever Presses"
  ) +
  scale_fill_manual(
    name = "Treatment",
    values = c("cno" = COLORS$cno, "veh" = COLORS$vehicle)
  ) +
  scale_color_manual(
    name = "Sex",
    values = c("male" = COLORS$male, "female" = COLORS$female)
  ) +
  stat_compare_means(
    method = "wilcox.test",
    paired = TRUE,
    label = "p.signif",
    size = 7
  ) +
  facet_wrap(~group * day) +
  theme_zi()
```



## 3. Fiber photomtery. Reward Discrimination Learning

```{r fiber-reward-discrimination}

# raw files /Users/laura/Downloads/zi-opto-fiber/mixed_fiber/TH-vGAT/rewarded_non_rewarded_lights
# png /Users/laura/Downloads/zi-opto-fiber/fiber_photometry_output
#python 
# Load reward/light discrimination data
light_reward_picked = read.csv("data/opto_light_rewarded.csv")

light_reward_picked$training = factor(light_reward_picked$training, levels = c("before", "after"))


#  reward discrimination
ggplot(light_reward_picked,aes(x = signal, y = as.numeric(auc))
) +
    geom_boxplot(aes(fill = signal), size = 0.8, alpha = 0.5) +
    geom_line(aes(group = filename), size = 0.5, alpha = 0.9) +
    geom_point(size = 3) +
    labs(
        title = "ZI Discriminates Rewarded Light After Training",
        x = "Light Type",
        y = "A.U.C"
    ) +
    scale_fill_manual(
        name = "Light",
        values = c("non_rewarded" = COLORS$no_reward, "rewarded" = COLORS$rewarded)
    ) +
    scale_color_manual(
        name = "Light",
        values = c("non_rewarded" = COLORS$no_reward, "rewarded" = COLORS$rewarded)
    ) +
    stat_compare_means(
        method = "wilcox.test",
        paired = TRUE,
        size = 6, label = "p.signif"
    ) +
    facet_wrap(~training) +
    theme_zi()

## too match python hist 
r_rewarded_light = ggplot(light_reward_picked,aes(x = signal, y = as.numeric(auc))
) +
    geom_boxplot(aes(fill = signal), size = 0.8, alpha = 0.8) +
    geom_line(aes(group = filename), size = 0.5, alpha = 0.3) +
    geom_point(size = 2) +
    labs(
        #title = "ZI Discriminates Rewarded Light After Training",
        #x = "Light Type",
        y = "A.U.C"
    ) +
    scale_fill_manual(
        name = "Light",
        values = c("non_rewarded" =	"#FFC896", "rewarded" = "#BBD6E8")
    ) +
    scale_color_manual(
        name = "Light",
        values = c("non_rewarded" =	"#FFC896", "rewarded" = "#BBD6E8")
    ) +
    stat_compare_means(
        method = "wilcox.test",
        paired = TRUE,
        size = 6, label = "p.signif"
    ) +
    facet_wrap(~training) +
    theme_zi()

ggsave("/Users/laura/Desktop/GRADSCHOOL/DIAS/manuscripts_grants/ZI_MANUSCRIPT/FIGURES/r_rewarded_light.pdf", plot = r_rewarded_light, width = 10.5, 
       height=6.5,  units = "in", device = cairo_pdf, dpi = 300)


wilcox.test(light_reward_picked[light_reward_picked$signal=="rewarded" & light_reward_picked$training=="before",]$auc,light_reward_picked[light_reward_picked$signal=="non_rewarded" & light_reward_picked$training=="before",]$auc)

wilcox.test(light_reward_picked[light_reward_picked$signal=="rewarded" & light_reward_picked$training=="after",]$auc,light_reward_picked[light_reward_picked$signal=="non_rewarded" & light_reward_picked$training=="after",]$auc)

reward <- light_reward_picked[light_reward_picked$signal=="rewarded" & light_reward_picked$training=="after",]$auc

non_reward <- light_reward_picked[light_reward_picked$signal=="non_rewarded" & light_reward_picked$training=="after",]$auc


cohen.d(reward, non_reward)


anova_lights <- aov(
  auc ~ signal*training*sex,
  data = light_reward_picked)

summary(anova_lights)

#sample size
table(light_reward_picked[light_reward_picked$training=="before" & light_reward_picked$signal=="rewarded",]$sex)


### training progress on light 
light_rewarded_not_training = read.csv("data/opto_training_progress.csv")

light_rewarded_not_training = light_rewarded_not_training %>% dplyr::group_by(Subject, Day) %>%
        dplyr::summarise(IRT_light_active = mean(IRT_light_active), IRT_light_passive = mean(IRT_light_passive))


light_rewarded_not_training_long <- light_rewarded_not_training %>%                  pivot_longer(
      cols = c(IRT_light_active, IRT_light_passive),                            
      names_to = "light",
      values_to = "IRT"
    ) %>%
    mutate(
      light = case_when(
        light == "IRT_light_active" ~ "rewarded",
        light == "IRT_light_passive" ~ "non-rewarded"
      )
    )
 

library(lmerTest)  

model <- lmer(IRT ~ light * Day + (1 | Subject), data =
                  light_rewarded_not_training_long)
results <- anova(model)                                                     
                                                                                
  # Get interaction p-value                                                     
  p_interaction <- results["light:Day", "Pr(>F)"]                               

  # Format as stars
  p_star <- ifelse(p_interaction < 0.001, "***",
            ifelse(p_interaction < 0.01, "**",
            ifelse(p_interaction < 0.05, "*", "ns")))
  
  
fiber_training_r = ggplot(light_rewarded_not_training_long[light_rewarded_not_training_long$Day<15 & light_rewarded_not_training_long$Subject %!in% c("th1.1", "th1.2", "th1.3", "h64.1", "h64.3"),],
       aes(x = Day, y = IRT, color = light)) +
    # mean points per group
    stat_summary(aes(group = light),
                 fun = mean,
                 geom = "point",
                 size = 2) +
    # mean ± SE per group
    stat_summary(aes(group = light),
                 fun.data = mean_se,
                 geom = "errorbar",
                 width = 0.3) +
    # mean line per group
    stat_summary(aes(group = light),
                 fun = mean,
                 geom = "line",
                 size = 1.2) +
    labs(
        title = "Training progress",
        x = "Training day",
        y = "IRT: light - detector, sec"
    ) +
    theme_zi() +
    theme(axis.text.x = element_text(angle = 0, hjust = 1)) +
    scale_color_manual(
        name = "Light",
        values = c("non-rewarded" =	"#FFC896", "rewarded" = "#BBD6E8")
    )+    annotate("text",
             x = 7, y = 27,
             label = paste0("light × day: ", p_star),
             size = 5)


ggsave("/Users/laura/Desktop/GRADSCHOOL/DIAS/manuscripts_grants/ZI_MANUSCRIPT/FIGURES/R_fiber_training.pdf", plot = fiber_training_r, width = 9.5, 
       height=5.5,  units = "in", device = cairo_pdf, dpi = 300)


```


---

# Optogenetics: Direct ZI Manipulation

## Optogenetic Stimulation During PR

```{r opto-pr-analysis}
# Load optogenetic PR data
all_opto_pr <- read.csv("data/pr_opto.csv")
all_opto_pr$stim <- factor(
  all_opto_pr$stim,
  levels = c("no_stim", "light_stim", "random_stim")
)

stim_comp <- list(
  c("no_stim", "light_stim"),
  c("no_stim", "random_stim"),
  c("light_stim", "random_stim")
)

day_comp<- list(
  c("d1", "d2"),
  c("d1", "d3"),
  c("d2", "d3")
)


all_opto_pr$stim = factor(all_opto_pr$stim, levels = c("no_stim", "random_stim", "light_stim"))

# Main optogenetic PR plot by sex
r_opto_pr = ggplot(
  all_opto_pr[all_opto_pr$animal %!in% c("h56.2", "h168.1") &
              all_opto_pr$freq == "20hz" &
              all_opto_pr$stim %in% c("light_stim", "random_stim", "no_stim"),],
  aes(x = stim, y = lev_press)
) +
  geom_boxplot(aes(fill = stim), alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.1, color = "black", size = 1.5, alpha = 0.6) +
  geom_line(aes(group = animal), alpha = 0.3) +
  labs(
    #title = "Optogenetic Stimulation During # PR: Sex Differences",
    x = "Stimulation Condition",
    y = "Lever Presses"
  ) +
  scale_fill_manual(
    name = "Stimulation",
    values = c(
      "no_stim" = COLORS$no_stim,
      "light_stim" = COLORS$light_stim,
      "random_stim" = COLORS$random_stim
    )
  ) +
  stat_compare_means(
    method = "wilcox.test",
    paired = TRUE,
    comparisons = stim_comp,
    label = "p.signif",
    size = 7
  ) +
  facet_wrap(~sex) +
  theme_zi()


ggsave("/Users/laura/Desktop/GRADSCHOOL/DIAS/manuscripts_grants/ZI_MANUSCRIPT/FIGURES/r_opto_pr.pdf", plot = r_opto_pr, width = 11.5, 
       height=5.5,  units = "in", device = cairo_pdf, dpi = 300)


# Main optogenetic PR plot by sex by day
r_opto_pr_days = ggplot(
  all_opto_pr[all_opto_pr$animal %!in% c("h56.2", "h168.1") &
              all_opto_pr$freq == "20hz" &
              all_opto_pr$stim %in% c("light_stim", "random_stim", "no_stim"),],
  aes(x = day, y = lev_press)
) +
  geom_boxplot(aes(fill = day), alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.1, color = "black", size = 2.5, alpha = 0.6) +
  geom_line(aes(group = animal), alpha = 0.3) +
  labs(
    title = "Optogenetic Stimulation During PR: Sex Differences",
    x = "Day",
    y = "Lever Presses"
  ) +
  scale_fill_manual(
    name = "Stimulation",
    values = c(
      "no_stim" = COLORS$no_stim,
      "light_stim" = COLORS$light_stim,
      "random_stim" = COLORS$random_stim
    )
  ) +
  stat_compare_means(
    method = "wilcox.test",
    paired = TRUE,
    comparisons = day_comp,
    label = "p.signif",
    size = 7
  ) +
  facet_wrap(~sex) +
  theme_zi()


ggsave("/Users/laura/Desktop/GRADSCHOOL/DIAS/manuscripts_grants/ZI_MANUSCRIPT/FIGURES/r_opto_pr_days.pdf", plot = r_opto_pr_days, width = 11.5, 
       height=5.5,  units = "in", device = cairo_pdf, dpi = 300)

wilcox.test(all_opto_pr[all_opto_pr$animal %!in% c("h56.2", "h168.1") &
                     all_opto_pr$freq == "20hz" &  all_opto_pr$sex=="fem" & all_opto_pr$stim=="no_stim",]$lev_press, all_opto_pr[all_opto_pr$animal %!in% c("h56.2", "h168.1") &
                     all_opto_pr$freq == "20hz" &  all_opto_pr$sex=="fem" & all_opto_pr$stim=="light_stim",]$lev_press, paired=TRUE)

wilcox.test(all_opto_pr[all_opto_pr$animal %!in% c("h56.2", "h168.1") &
                            all_opto_pr$freq == "20hz" &  all_opto_pr$sex=="male" & all_opto_pr$stim=="light_stim",]$lev_press, all_opto_pr[all_opto_pr$animal %!in% c("h56.2", "h168.1") &
                                                                                                                                            all_opto_pr$freq == "20hz" &  all_opto_pr$sex=="male" & all_opto_pr$stim=="no_stim",]$lev_press, paired=TRUE)


opto_anova = aov(lev_press~sex:stim, data = all_opto_pr[all_opto_pr$animal %!in% c("h56.2", "h168.1") & all_opto_pr$freq == "20hz",])

summary(opto_anova)

```

## 5.2 Training Progress Before Optogenetic Testing

```{r opto-training-progress}
# Load training data
all_opto_training_sum <- read.csv("data/opto_training.csv")
all_opto_meta <- read.csv("data/opto_meta.csv")
all_opto_training_sum <- merge(all_opto_training_sum, all_opto_meta)


library(lme4)                               
  library(lmerTest)                       
                                              
  model <- lmer(pressing_rate ~ sex * day + (1 | animal),                       
                data = all_opto_training_sum[all_opto_training_sum$excl ==
  "incl", ])                                                                    
  results <- anova(model)
                                                                                
  # Format interaction p-value                                                  
  p_int <- results["sex:day", "Pr(>F)"]
  p_star <- ifelse(p_int < 0.001, "***",
            ifelse(p_int < 0.01, "**",
            ifelse(p_int < 0.05, "*", "ns")))
  
  
# Press rate across training by sex
ggplot(all_opto_training_sum[all_opto_training_sum$excl == "incl",],
  aes(x = day, y = pressing_rate, color = sex)) +
  stat_summary(fun = "mean", geom = "line", aes(group = sex), size = 1.2) +
  stat_summary(fun = "mean", geom = "point", size = 2) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.3) +
  labs(
    title = "Training Progress: Press Rate by Sex",
    x = "Training Day",
    y = "Press Rate (presses/min)"
  ) +
  scale_color_manual(
    name = "sex",
    values = c("male" = COLORS$male, "female" = COLORS$female)) +
  theme_zi()+annotate("text", x = 3, y = Inf,
             label = paste0("sex × day: ", p_star),
             vjust = 1.5, size = 4)

#average pressing across sexes

library(rstatix, lib.loc = "/Library/Frameworks/R.framework/Versions/4.1/Resources/library")

stat_overall <- all_opto_training_sum %>%
    filter(excl == "incl") %>%
    dplyr::group_by(sex, animal, state) %>%
    dplyr::summarise(mean_rate = mean(pressing_rate, na.rm = TRUE)) %>%
    dplyr::group_by(sex) %>%
    wilcox_test(mean_rate ~ state, paired = TRUE) %>% add_significance()

stat_labels <- stat_overall %>%
    mutate(  x = 4,    y = Inf,label = paste0(p.signif))  

ggplot(data = all_opto_training_sum[all_opto_training_sum$excl=="incl",], aes(x = day, y = pressing_rate, color = state)) +
    stat_summary(fun = "mean", geom = "line", aes(group=as.factor(state)), size = 1) +
    stat_summary(fun = "mean", geom = "point",
                 size = 1) +
    stat_summary(fun.data = "mean_se",
                 geom = "errorbar",
                 width = 0.3) +labs(title = "", x="day", y = "pressing rate", color = "training block") + theme(text = element_text(size = 14), axis.text.x = element_text(face="bold", angle=45))+scale_color_manual(values = c("brown", "black"), labels = c("light on, lever active", "light off, lever not active"))+
  theme_zi()+facet_wrap(.~sex)+
    geom_text(data = stat_labels,
              aes(x = x, y = y, label = label),
              vjust = 1.5, size = 4,
              inherit.aes = FALSE)

```

## 5.3 Inter-Response Time (IRT) Analysis

```{r opto-irt-analysis}
# Load IRT data
all_cohorts_irt_sum_filt2 <- read.csv("data/opto_irt.csv")
opto_meta <- read.csv("data/opto_meta.csv")
all_cohorts_irt_sum_filt2 <- merge(all_cohorts_irt_sum_filt2, opto_meta)
#success rate 

opto_success <- read.csv("data/opto_success.csv")
opto_success = merge(opto_success, opto_meta)

ggplot(
  opto_success,
  aes(x = day, y = succes_rate)
) +
  stat_summary(
    fun = "mean",
    geom = "line",
    aes(group = sex, color = sex),
    size = 1.2
  ) +
  stat_summary(
    fun = "mean",
    geom = "point",
    aes(color = sex),
    size = 2
  ) +
  stat_summary(
    fun.data = "mean_se",
    geom = "errorbar",
    aes(color = sex),
    width = 0.3
  ) +
  labs(
    title = "Succcess Rate During Light Press Training",
    x = "Training Day",
    y = "rate"
  ) +
  scale_color_manual(
    name = "Sex",
    values = c("male" = COLORS$male, "female" = COLORS$female)
  ) +
  theme_zi()

model_success <- lmer(succes_rate ~ sex * day + (1 | animal), data =
  opto_success)                           
  results_success <- anova(model_success)
    
  results_success
  
  p_int <- results_success["sex:day", "Pr(>F)"]
  p_star <- ifelse(p_int < 0.001, "***",                                        
            ifelse(p_int < 0.01, "**",
            ifelse(p_int < 0.05, "*", "ns")))

  

# IRT across training by sex
ggplot(
  all_cohorts_irt_sum_filt2[all_cohorts_irt_sum_filt2$incl == "incl",],
  aes(x = day, y = irt)
) +
  stat_summary(
    fun = "mean",
    geom = "line",
    aes(group = sex, color = sex),
    size = 1.2
  ) +
  stat_summary(
    fun = "mean",
    geom = "point",
    aes(color = sex),
    size = 2
  ) +
  stat_summary(
    fun.data = "mean_se",
    geom = "errorbar",
    aes(color = sex),
    width = 0.3
  ) +
  labs(
    title = "Inter-Response Time During Light Press Training",
    x = "Training Day",
    y = "IRT (s)"
  ) +
  scale_color_manual(
    name = "Sex",
    values = c("male" = COLORS$male, "female" = COLORS$female)
  ) +
  theme_zi()

##showing the difference in training 

library(tidyverse)
library(broom)

# Calculate means and differences with statistics
irt_stats <- all_cohorts_irt_sum_filt2 %>%
  dplyr::filter(incl == "incl") %>%
  dplyr::group_by(day) %>%
  tidyr::nest() %>%
  dplyr::mutate(
    # Calculate means by sex
    means = purrr::map(data, ~{
      .x %>%
        dplyr::group_by(sex) %>%
        dplyr::summarise(
          mean_irt = mean(irt, na.rm = TRUE),
          se_irt = sd(irt, na.rm = TRUE) / sqrt(n()),
          .groups = "drop"
        )
    }),
    # Perform t-test
    t_test = purrr::map(data, ~{
      if(length(unique(.x$sex)) == 2) {
        t.test(irt ~ sex, data = .x) %>% broom::tidy()
      } else {
        tibble(p.value = NA, statistic = NA)
      }
    })
  ) %>%
  tidyr::unnest(c(means, t_test)) %>%
  tidyr::pivot_wider(
    names_from = sex,
    values_from = c(mean_irt, se_irt)
  ) %>%
  dplyr::mutate(
    diff = mean_irt_male - mean_irt_female,
    diff_se = sqrt(se_irt_male^2 + se_irt_female^2),
    significant = p.value < 0.05
  )

# Plot with statistical annotations
ggplot(irt_stats, aes(x = day, y = diff)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", size = 0.8) +
  geom_line(size = 1.2, color = "black") +
  geom_point(
    aes(shape = significant, fill = significant),
    size = 3,
    color = "black"
  ) +
  geom_errorbar(
    aes(ymin = diff - diff_se, ymax = diff + diff_se),
    width = 0.3
  ) +
  scale_shape_manual(
    name = "p < 0.05",
    values = c("TRUE" = 21, "FALSE" = 21)
  ) +
  scale_fill_manual(
    name = "p < 0.05",
    values = c("TRUE" = "black", "FALSE" = "white")
  ) +
  labs(
    title = "Sex Difference in IRT During Training",
    subtitle = "Male - Female",
    x = "Training Day",
    y = "IRT Difference (s)"
  ) +
  theme_zi()

# Print summary statistics table
irt_stats %>%
  dplyr::select(day, diff, diff_se, statistic, p.value) %>%
  dplyr::mutate(
    p.value = format.pval(p.value, digits = 3),
    dplyr::across(c(diff, diff_se, statistic), ~round(.x, 3))
  ) %>%
  print()


```

**Key Finding:** Optogenetic results reveal causal relationship between sensory stimuli and motivational drive 

---

# 6. Food Consumption (GI DREADD)

```{r food-consumption-analysis}
# Load food consumption data
h172h175_pellets <- read.csv("data/pit_pellets_probe.csv")
pit_gi_meta <- read.csv("data/pit_meta_gi.csv")
h172h175_pellets <- merge(h172h175_pellets, pit_gi_meta)

table(pit_gi_meta$sex, pit_gi_meta$group)

h172h175_pellets$food_restriction <- factor(
  h172h175_pellets$food_restriction,
  levels = c("long", "acute")
)

table(h172h175_pellets[h172h175_pellets$food_restriction=="long",]$sex, h172h175_pellets[h172h175_pellets$food_restriction=="long",]$virus)

# Food consumption plot
ggplot(h172h175_pellets[h172h175_pellets$food_restriction=="long",], aes(x = virus, y = as.numeric(pellets))) +
  geom_boxplot(aes(fill = virus), size = 1, alpha = 0.5, outlier.shape = NA) +
  geom_point(size = 3) +
  labs(
    title = "Food Consumption: GI DREADD with CNO",
    x = "Group",
    y = "Pellets Consumed"
  ) +
  scale_fill_manual(
    name = "Virus",
    values = c("gi" = COLORS$gi_dreadd, "gfp" = COLORS$gfp)
  ) +
  scale_color_manual(
    name = "Sex",
    values = c("male" = COLORS$male, "female" = COLORS$female)
  ) +
  stat_compare_means(
    method = "wilcox.test",
    paired = FALSE,
    size = 7,
    label = "p.format"
  ) +
  theme_zi()

food_aov = aov(pellets ~ virus, data = h172h175_pellets[h172h175_pellets$food_restriction=="long",])
summary(food_aov)

wilcox.test(h172h175_pellets[h172h175_pellets$food_restriction=="long" & h172h175_pellets$virus=="GFP",]$pellets, h172h175_pellets[h172h175_pellets$food_restriction=="long" & h172h175_pellets$virus=="GI",]$pellets)


```

**Key Finding:** GI manipulation affects food consumption .

---

# 7. Stress and Motivation (GQ DREADD)

```{r stress-gq-analysis}
# Load stress GQ data
stress_gq_pr <- read.csv("data/pr_stress_gq.csv")
breakpoint_stress <- read.csv("data/breakpoint_stress_gq.csv")
stress_gq_pr <- merge(stress_gq_pr, breakpoint_stress)
stress_gq_pr$treatm <- factor(stress_gq_pr$treatm, levels = c("veh", "cno"))

cno_group_comp <- list(
  c("cont", "stress_gfp"),
  c("cont", "stress_gq"),
  c("stress_gfp", "stress_gq")
)


# Between-group comparison (CNO day, susceptible animals only)
stress_gq_pr = stress_gq_pr[stress_gq_pr$animal %!in% c("S6.3", "S5.3"),]

stress_ggplot = ggplot(
    stress_gq_pr[stress_gq_pr$animal %!in% c("S6.3", "S5.3"),],
    aes(x = group, y = breakingpoint)
) +
    geom_boxplot(aes(fill = group), alpha = 0.7, outlier.shape = NA) +
    geom_jitter(aes(color = sex), width = 0.1, size = 2.5, alpha = 0.6) +
    geom_line(aes(group = animal), alpha = 0.3) +
    labs(
        #title = "Stress Model: ZI Activation Rescues Motivation",
        x = "Group",
        y = "Breakpoint"
    ) +
    scale_fill_manual(
        name = "Group",
        values = c(
            "cont" = COLORS$control,
            "stress_gfp" = COLORS$stress,
            "stress_gq" = COLORS$gq_dreadd
        )
    ) +
    scale_color_manual(
        name = "Sex",
        values = c("male" = COLORS$male, "female" = COLORS$female)
    ) +
    stat_compare_means(
        method = "wilcox.test",
        paired = FALSE,
        size = 7,
        comparisons = cno_group_comp, label = "p.signif"
    )+ facet_wrap(~treatm)+theme_zi()


ggsave("/Users/laura/Desktop/GRADSCHOOL/DIAS/manuscripts_grants/ZI_MANUSCRIPT/FIGURES/R_gq_stress.pdf", plot = stress_ggplot, width = 11.5, 
       height=5,  units = "in", device = cairo_pdf, dpi = 300)


anova_stress = aov(breakingpoint ~ group*treatm*sex, data =  stress_gq_pr[stress_gq_pr$animal %!in% c("S6.3", "S5.3"),])
summary(anova_stress)

#mixed model for paper

model_stress_pr <- lmer(breakingpoint ~ treatm * group * sex + (1 | animal), data =stress_gq_pr[stress_gq_pr$animal %!in% c("S6.3", "S5.3"),])
anova(model_stress_pr)



wilcox.test(stress_gq_pr[stress_gq_pr$animal %!in% c("S6.3", "S5.3") & stress_gq_pr$group =="cont" & stress_gq_pr$treatm=="veh",]$breakingpoint, stress_gq_pr[stress_gq_pr$animal %!in% c("S6.3", "S5.3") & stress_gq_pr$group =="stress_gfp" & stress_gq_pr$treatm=="veh",]$breakingpoint)

cont <- stress_gq_pr[stress_gq_pr$group=="cont" & stress_gq_pr$treatm=="veh" ,]$breakingpoint

stress_gq <- stress_gq_pr[stress_gq_pr$group=="stress_gq" & stress_gq_pr$treatm=="veh" ,]$breakingpoint

stress_gfp <- stress_gq_pr[stress_gq_pr$group=="stress_gfp" & stress_gq_pr$treatm=="veh" ,]$breakingpoint

cohen.d(cont, stress_gfp)

###FR1 training 

stress_meta = stress_gq_pr %>%
    dplyr::distinct(animal, group, sex)

stress_FR1_training = read.csv("data/stress_fr1_training.csv")

stress_FR1_training = merge(stress_FR1_training, stress_meta)

ggplot(stress_FR1_training,
       aes(x = day, y = rewards, color = group)) +
    # mean points per group
    stat_summary(aes(group = group),
                 fun = mean,
                 geom = "point",
                 size = 2) +
    
    # mean ± SE per group
    stat_summary(aes(group = group),
                 fun.data = mean_se,
                 geom = "errorbar",
                 width = 0.3) +
    
    # mean line per group
    stat_summary(aes(group = group),
                 fun = mean,
                 geom = "line",
                 size = 1.2) +
    
    labs(
        title = "Training progress",
        x = "Training day",
        y = "Reinforcements"
    ) +
    theme_zi() +
    theme(axis.text.x = element_text(angle = 0, hjust = 1)) +
    facet_wrap(~sex)+scale_color_manual(
        name = "Group",
        values = c(
            "cont" = COLORS$control,
            "stress_gfp" = COLORS$stress,
            "stress_gq" = COLORS$gq_dreadd
        )
    )
```

**Key Finding:** Stress reduces motivation, and ZI activation can rescue this deficit.

---


## Statistical Summary Tables

### 1. **Effect Size Calculations**
```{r}
# Calculate Cohen's d for main comparisons
library(effsize)

# For GI DREADD PR test
gi_effect <- cohen.d(
  breakingpoint ~ treatment,
  data = pr_h80h90[pr_h80h90$group == "GI",]
)
gi_effect

# For GQ DREADD PR test
gq_effect <- cohen.d(
  breakingpoint ~ treatment,
  data = hcohorts_gq_pr[hcohorts_gq_pr$group == "hm3dgq",]
)
gq_effect
```

---

### 2. **Learning Curve Analysis**

```{r}
# Analyze acquisition of operant behavior
learning_curves <- all_opto_training_sum %>%
  filter(excl == "incl") %>%
  group_by(day, sex) %>%
  summarise(
    mean_press = mean(pressing_rate),
    se_press = sd(pressing_rate) / sqrt(n())
  )
```
---

### 3. **Response Variability Analysis**

```{r}
# Calculate coefficient of variation (CV) for each group
cv_analysis <- pr_h80h90 %>%
  group_by(group, treatment) %>%
  summarise(
    mean_bp = mean(breakingpoint),
    sd_bp = sd(breakingpoint),
    cv = (sd_bp / mean_bp) * 100,
    n = n())

# Test for differences in variability (Levene's test)
library(car)
leveneTest(breakingpoint ~ group * treatment, data = pr_h80h90)
```


```{r} 
#PIT Sound
pit_sound_probe = read.csv("data/pit_sound_probe.csv")
pit_sound_probe$block = factor(pit_sound_probe$block, levels = c("beginning", "sound1", "light1", "sound2", "light2"))
pit_sound_probe$treatm = factor(pit_sound_probe$treatm, levels = c("veh", "cno"))
pit_sound_probe$group = factor(pit_sound_probe$group, levels = c("gfp", "gi"))
sex_colors <- c("male" = "#4292C6", "female" = "#E34A33")
block_comp = list(
  c("beginning", "sound1"),
  c("sound1", "light1"),
  c("beginning", "light1"), c("light1", "sound2"), c("sound1", "sound2"), c("light2", "sound2"), c("light1", "light2"))

#removed not sign comp 
block_comp = list(
  c("beginning", "sound1"),
  c("sound1", "light1"),
  c("beginning", "light1"), c("light1", "sound2"), c("light2", "sound2"))



ggplot(pit_sound_probe[pit_sound_probe$animal!="h189.3",], aes(x = block, y = rate)) +
    # Add boxplot
    geom_boxplot(aes(fill = block2), alpha = 0.5, width = 0.5) +
    # Add individual points colored by sex
    geom_jitter(aes(color = sex), width = 0.15, size = 2, alpha = 0.7) +
    scale_fill_manual(values = c("#CCCCCC", "lightblue", "lightyellow", "lightblue", "lightyellow")) +
    scale_color_manual(name = "Sex", values = c("male" = COLORS$male, "female" = COLORS$fem)) +
    labs(title = "probe test after PIT. paired comparisons",
         x = "Block",
         y = "Pressing Rate") +
    theme_classic() +
    theme_zi()+facet_wrap(.~group*treatm)+stat_compare_means(
        method = "wilcox.test",
        label = "p.signif",
        paired = TRUE,
        size = 5, comparisons = block_comp)

#CNO ONLY 

ggplot(pit_sound_probe[pit_sound_probe$animal!="h189.3" & pit_sound_probe$treatm=="cno",], aes(x = block, y = rate)) +
    # Add boxplot
    geom_boxplot(aes(fill = block2), alpha = 0.5, width = 0.5) +
    # Add individual points colored by sex
    geom_jitter(aes(color = sex), width = 0.15, size = 1.7, alpha = 0.9) +
    scale_fill_manual(values = c("#CCCCCC",  "lightyellow", "lightblue","lightyellow", "lightblue")) +
    scale_color_manual(name = "Sex", values = c("male" = COLORS$male, "female" = COLORS$fem)) +
    labs(title = "probe test after PIT. paired comparisons",
         x = "Block",
         y = "Pressing Rate") +
    theme_classic() +
    theme_zi()+facet_wrap(.~group)+stat_compare_means(
        method = "wilcox.test",
        label = "p.signif",
        paired = TRUE,
        size = 5, comparisons = block_comp)+stat_summary(aes(group = 1),         fun = mean, geom = "line",                     linewidth = 1.2,                              color = "black",                         alpha = 0.7  )  

#sample size
table(pit_sound_probe[pit_sound_probe$animal!="h189.3" & pit_sound_probe$treatm=="cno" & pit_sound_probe$block2=="beginning",]$group, pit_sound_probe[pit_sound_probe$animal!="h189.3" & pit_sound_probe$treatm=="cno" & pit_sound_probe$block2=="beginning",]$sex)


##averaged across two light and two sounds 
pit_sound_probe_av = pit_sound_probe[pit_sound_probe$animal!="h189.3",] %>%
  dplyr::group_by(animal, sex, treatm,group, block2) %>%
  dplyr::summarise(rate = mean(rate))

block_comp_av = list(
  c("beginning", "sound"),
  c("beginning", "light"),
  c("sound", "light"))

pit_sound_probe_av$block2 = factor(pit_sound_probe_av$block2, levels = c("beginning", "sound","light"))

ggplot(pit_sound_probe_av, aes(x = block2, y = rate)) +
    # Add boxplot
    geom_boxplot(aes(fill = block2), alpha = 0.5, width = 0.5) +
    # Add individual points colored by sex
    geom_jitter(aes(color = sex), width = 0.15, size = 2.5, alpha = 0.7) +
    scale_fill_manual(values = c("#CCCCCC", "lightblue", "lightyellow")) +
    scale_color_manual(values = sex_colors, name = "Sex") +
    labs(title = "probe test after PIT. paired comparisons",
         x = "Block",
         y = "Pressing Rate") +
    theme_classic() +
    theme(
        plot.title = element_text(size = 10, face = "bold", hjust = 0.5),
        axis.title = element_text(size = 9),
        axis.text = element_text(size = 8),
        legend.position = "right",
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 7),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
    ) +facet_wrap(.~group*treatm)+stat_compare_means(
        method = "wilcox.test",
        paired = TRUE,
        size = 7, comparisons = block_comp_av)


pit_anova = aov(rate ~ block*sex*group, data = pit_sound_probe[pit_sound_probe$treatm=="cno",])
summary(pit_anova)

library(lme4)
  library(lmerTest)                                                             
   
  model <- lmer(rate ~ block * sex * group + (1 | animal),                     
                data = pit_sound_probe[pit_sound_probe$treatm == "cno", ])
  anova(model)

pit_anova2 = aov(rate ~ block2*sex*treatm*group, data = pit_sound_probe_av)
summary(pit_anova2)

wilcox.test(pit_sound_probe_av[pit_sound_probe_av$block2=="sound" & pit_sound_probe_av$treatm=="cno" & pit_sound_probe_av$group=="gi",]$rate, pit_sound_probe_av[pit_sound_probe_av$block2=="beginning" & pit_sound_probe_av$treatm=="cno" & pit_sound_probe_av$group=="gi",]$rate )
```


---

# Subjects Summary

```{r subjects-summary}
# Helper: counts unique subjects by sex and (optionally) by group
summarize_exp <- function(df, id_col, sex_col = "sex", group_col = NULL) {
  keep <- c(id_col, sex_col)
  if (!is.null(group_col) && group_col %in% names(df)) keep <- c(keep, group_col)
  d <- unique(df[, keep, drop = FALSE])
  d <- d[!is.na(d[[sex_col]]) & d[[sex_col]] != "", ]
  s <- tolower(substr(d[[sex_col]], 1, 1))
  fem <- sum(s == "f"); mal <- sum(s == "m")
  groups_str <- "—"; n_groups <- NA_integer_
  if (!is.null(group_col) && group_col %in% names(d)) {
    tab <- table(d[[group_col]])
    tab <- tab[tab > 0]
    n_groups <- length(tab)
    groups_str <- paste(sprintf("%s (n=%d)", names(tab), as.integer(tab)),
                        collapse = ", ")
  }
  list(female = fem, male = mal, n_groups = n_groups, groups = groups_str)
}

experiments <- list()

# GI PR: pre-filtered at load; plots/stats use cohort %in% c("h90","h80")
pr_gi_f <- pr_h80h90[pr_h80h90$cohort %in% c("h90", "h80"), ]
experiments[["GI DREADD - Progressive Ratio"]] <-
  summarize_exp(pr_gi_f, "Animal", group_col = "group")

# GI Open Field: pre-filtered at load; plots/stats add excl!="excl" & Distance<15000
of_raw <- open_field_h80h90[open_field_h80h90$excl != "excl" &
                            open_field_h80h90$Distance < 15000, ]
of <- merge(unique(of_raw[, "Animal", drop = FALSE]), meta, by = "Animal")
experiments[["GI DREADD - Open Field"]] <- summarize_exp(of, "Animal", group_col = "group")

# GI Memory: no additional filter
experiments[["GI DREADD - Memory"]] <- summarize_exp(h90_memory, "Animal", group_col = "group")

# GQ PR: no filter in primary plot
experiments[["GQ DREADD - Progressive Ratio"]] <-
  summarize_exp(hcohorts_gq_pr, "animal", group_col = "group")

# GQ Probe: animal %!in% c("18.2","H112.2") & cort != "cort"
gqp_raw <- h_gq_prob_test[h_gq_prob_test$animal %!in% c("18.2", "H112.2"), ]
gqp <- merge(unique(gqp_raw[, "animal", drop = FALSE]), h_meta, by = "animal")
gqp <- gqp[gqp$cort != "cort", ]
experiments[["GQ DREADD - Probe Test"]] <- summarize_exp(gqp, "animal", group_col = "group")

# Fiber photometry: no filter
experiments[["Fiber Photometry - Reward Discrimination"]] <-
  summarize_exp(light_reward_picked, "filename")

# Opto PR: animal %!in% c("h56.2","h168.1") & freq == "20hz"
opto_pr_f <- all_opto_pr[all_opto_pr$animal %!in% c("h56.2", "h168.1") &
                         all_opto_pr$freq == "20hz", ]
experiments[["Optogenetic - Progressive Ratio"]] <-
  summarize_exp(opto_pr_f, "animal")

# Opto Attention Training: excl == "incl"
ot_f <- all_opto_training_sum[all_opto_training_sum$excl == "incl", ]
experiments[["Optogenetic - Attention Training"]] <- summarize_exp(ot_f, "animal")

# Opto IRT: incl == "incl"
oi_f <- all_cohorts_irt_sum_filt2[all_cohorts_irt_sum_filt2$incl == "incl", ]
experiments[["Optogenetic - IRT"]] <- summarize_exp(oi_f, "animal")

# PIT Pellets Probe: food_restriction == "long"
pitp <- h172h175_pellets[h172h175_pellets$food_restriction == "long", ]
experiments[["PIT - Pellets Probe"]] <- summarize_exp(pitp, "animal", group_col = "group")

# PIT Sound Probe: animal != "h189.3" & treatm == "cno"
pit_sp_f <- pit_sound_probe[pit_sound_probe$animal != "h189.3" &
                            pit_sound_probe$treatm == "cno", ]
experiments[["PIT - Sound Probe"]] <-
  summarize_exp(pit_sp_f, "animal", group_col = "group")

# Stress GQ PR: animal %!in% c("S6.3","S5.3")
stress_pr_f <- stress_gq_pr[stress_gq_pr$animal %!in% c("S6.3", "S5.3"), ]
experiments[["Stress GQ DREADD - Progressive Ratio"]] <-
  summarize_exp(stress_pr_f, "animal", group_col = "group")

subj_summary <- do.call(rbind, lapply(names(experiments), function(nm) {
  v <- experiments[[nm]]
  data.frame(
    Experiment = nm,
    Female = v$female,
    Male = v$male,
    Total = v$female + v$male,
    `# Groups` = v$n_groups,
    Groups = v$groups,
    check.names = FALSE
  )
}))

write.csv(subj_summary, "data/subjects_summary.csv", row.names = FALSE)

knitr::kable(
  subj_summary,
  caption = "Subject counts (female / male) and groups by experiment",
  row.names = FALSE
) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```
